---
title: "Teenage Pregnancy in Malawi: Risk Factor and Spatial Analysis"
author: "B.Banda"
date: "Malawi MICS 2019-2020"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: united
    code_folding: show
    df_print: paged
    keep_md: false
editor_options:
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)

# Set working directory so all relative paths (.shp, .sav, output figures)
# resolve to the MICS data folder regardless of where chunks are run.
knitr::opts_knit$set(root.dir = "C:/Users/b.banda/Box/Bernard_Banda_Files/MICS_Spatial_Analysis")
```

# PART I: Background and Methods

## Study Overview / Background

Adolescent pregnancy remains a pressing global public health concern,
with an estimated 16 million girls aged 15-19 giving birth each year,
the majority in low- and middle-income countries. In Malawi, teenage
pregnancy carries serious consequences for maternal and child health,
education, and socioeconomic mobility. Despite national efforts to
expand reproductive health services and adolescent-friendly programming,
the prevalence of teenage pregnancy remains high, and substantial
geographic variation persists across districts and sub-district
administrative units.

This analysis uses the 2019-2020 Malawi Multiple Indicator Cluster
Survey (MICS) to (i) quantify the prevalence of teenage pregnancy among
adolescent females (15-19 years) and identify socio-demographic and
economic risk factors using survey-weighted logistic regression, and
(ii) characterise the spatial distribution of teenage pregnancy at the
Traditional Authority (TA) level using global and local indicators of
spatial autocorrelation to identify hotspots and coldspots.

## Methods

This study used data from the 2019-2020 Malawi MICS, a nationally
representative household survey conducted by the National Statistical
Office with support from UNICEF. The MICS employed a two-stage
stratified cluster sampling design: enumeration areas (clusters) were
selected within geographic strata (district x urbanicity combinations),
and households were then selected by systematic sampling within each
cluster. The survey collected data from 25,626 women aged 15-49 across
1,111 enumeration area clusters.

This analysis examined a subsample of adolescent females aged 15-19 (n =
5,770 respondents from 1,086 clusters across 333 Traditional
Authorities). Teenage pregnancy was defined as any female aged 15-19 who
reported either ever having given birth (CM1 = 1) or currently being
pregnant. Women with missing data on age, pregnancy status, or key
sociodemographic variables were excluded (n = 12; final analytical
sample: 5,758 women).

### Survey Design Specification

To properly account for the complex sampling structure, a survey design
object was constructed in R specifying the primary sampling unit (HH1,
the enumeration area cluster), strata (39 district x urbanicity
combinations), and individual woman weights (`wmweight`), with clusters
nested within strata (`nest = TRUE`). Individual weights ranged from
0.089 to 16.026 (mean = 2.01), indicating a non-self-weighting design
requiring statistical adjustment.

### Risk Factor Analysis

Sociodemographic and economic exposures included education level, wealth
index quintile, place of residence (urbanicity), current marital status,
health insurance coverage, and contraceptive use. Survey-weighted
prevalence of teenage pregnancy was estimated overall and stratified by
each exposure using `svyby()` with `svyciprop` (logit-based 95% CI).
Bivariate associations were tested using Pearson χ² applied to
survey-weighted contingency tables (`svytable()` + `chisq.test()`).
Survey-weighted logistic regression (`svyglm`, family = quasibinomial)
was used for both unadjusted (Model 1, separate univariable models) and
fully adjusted (Model 2, single multivariable model) specifications. 95%
confidence intervals were computed using Wald-type variance estimates
with the Taylor-series linearisation method implemented in the `survey`
package, accounting for the complex two-stage stratified cluster design.
Reference categories were set *a priori* using the most-protected or
most-prevalent stratum: Education = Primary or lower; Wealth Index =
Richest; Place of Residence = Urban; Marital status = No - not in union;
Contraceptive use = No; Health insurance = No. Item-level missingness
was low (\<2% on all covariates; see Section 6); the multivariable model
used a complete-case analytic sample with the unweighted and weighted
sample sizes reported in the Table 3 footer.

### Spatial Analysis

Individual pregnancy status and survey weights were linked to cluster
identifiers (HH1) and Traditional Authority (TA) names via the GPS
cluster dataset (`MalawiMICS2019-20GPS.shp`). After merging, 1,095
clusters mapped to 333 unique Traditional Authorities. To account for
GPS jittering applied in MICS for confidentiality (+/- 2-10 km urban,
+/- 5 km rural), prevalence estimates were aggregated to the TA level.
For each TA we computed the survey-weighted prevalence of pregnancy, 95%
confidence intervals, unweighted sample size, and count of pregnant
teens. TAs with fewer than 10 teen respondents were flagged as "very
small" and those with 10-19 respondents as "small," but all TAs with at
least one respondent were retained to avoid selection bias.

TA boundary polygons were obtained from `mics_boundaries_nr.shp`.
Polygons were aggregated to one geometry per unique TA name using
spatial union (TA Nkula was excluded due to topology errors; final n =
331 unique TAs in WGS84). Prevalence values were standardised using
Z-score transformation prior to spatial statistical testing.

Spatial relationships were defined using K-nearest neighbours (KNN-5)
based on TA centroids, with row-standardised weights (style = `"W"`).
Global Moran's I was tested using both the asymptotic test (assuming
normality) and a Monte Carlo permutation test (999 simulations). Local
Indicators of Spatial Association (LISA) were computed for each TA,
classifying each into High-High (Hotspot), Low-Low (Coldspot), High-Low
or Low-High (Spatial Outlier), or Not Significant categories based on
quadrant position and statistical significance (alpha = 0.05).

### Statistical Software

All analyses were performed in R (version 4.4.0). Survey-weighted
descriptive and inferential statistics were computed with the `survey`
package (Lumley, 2024) and `srvyr`. Data manipulation and tabulation
used `tidyverse`, `haven`, `janitor`, `scales`, `gtsummary`, and `gt`.
Spatial preprocessing, autocorrelation statistics, and mapping used
`sf`, `spdep`, `sfdep`, `ggplot2`, `patchwork`, and `ggspatial`.
Missingness was visualised with `mice::md.pattern()` (no imputation was
performed; see Section 6). All tables and plots are generated
reproducibly from this R Markdown document.

## Exposure Variables

The pre-specified exposures examined in the multivariable model (Table
3, Model 2) were:

-   **Education**: Primary or lower (reference) vs. Secondary or higher
-   **Wealth Index**: Richest (reference), Fourth, Middle, Second,
    Poorest (derived from household asset ownership and housing
    characteristics)
-   **Place of Residence**: Urban (reference) vs. Rural
-   **Marital status**: No - not in union (reference), Yes - currently
    married, Yes - living with partner
-   **Contraceptive use**: No (reference) vs. Yes (currently using any
    contraceptive method)
-   **Health insurance**: No (reference) vs. Yes

An additional exposure, *Ever had sex* (No vs. Yes), is reported in the
descriptive Table 2 but excluded from the multivariable model on
conceptual grounds: sexual debut sits on the causal pathway between
sociodemographic exposures and pregnancy and would induce overadjustment
bias if included alongside the upstream exposures. All exposures are
treated as categorical, with reference categories chosen *a priori*
based on the most-protected or modal stratum.

## Outcome Variable

The primary outcome is teenage pregnancy, defined among females aged
15-19 in the MICS Women's dataset as any respondent who reported having
ever given birth (CM1 = 1) OR currently being pregnant. This composite
outcome captures both past pregnancy experience and current pregnancy
status, reflecting the full burden of adolescent fertility in this age
group.

# PART II: Data Preparation

## 1. Load Required Packages

We used the following packages in our analysis to enable data
manipulation, survey-weighted estimation, publication-quality tables,
multiple imputation, spatial analysis, and visualization. The pacman
package manager automatically installs any missing packages and loads
them into the R environment.

```{r load-packages, message=FALSE, warning=FALSE}
# Load required packages using pacman for convenience
pacman::p_load(
  # Data manipulation and utilities
  data.table, tidyverse, purrr, haven, janitor, labelled,

  # Survey analysis and weighted estimation
  survey, srvyr,

  # Publication tables and formatting
  gtsummary, gt, kableExtra, knitr,

  # Spatial analysis and mapping
  sf, sfdep, spdep, leaflet, ggspatial, ggrepel,

  # Visualization
  ggplot2, patchwork,

  # Statistical methods
  mice,

  # Other utilities
  DiagrammeR
)
```

## 2. Read Women's Dataset and Select Variables

We imported the MICS 2019–20 Women's dataset from an SPSS (.sav) file
and restricted the sample to adolescent females aged 15–19, retaining
only the variables required for analysis: cluster identifier, household
number, birth history, pregnancy status, age, education, urbanicity,
marital status, health insurance, contraceptive use, sexual activity,
sampling weight, wealth index, and stratum.

```{r read-women-data, message=FALSE}
WM <- read_sav("wm.sav") %>%
  clean_names()

wm_small <- WM %>%
  filter(wb4 >= 15 & wb4 <= 19) %>%
  select(
    hh1, hh2, cm1, cp1, wb3m, wb3y, wb4, wb6a, wb14, wb16, wb17, wb18,
    cp3, ma11, sb1, wmweight, windex5, ma1, stratum
  )
```

## 3. Open MICS Shapefile and GIS Data

We loaded two spatial datasets: the GPS cluster shapefile containing
enumeration area coordinates (`MalawiMICS2019-20GPS.shp`), which links
each cluster to its geographic location and Traditional Authority name
(`GEONAMET`), and the TA boundary polygons (`mics_boundaries_nr.shp`)
used for choropleth mapping and spatial analysis.

```{r open-gis-data, message=FALSE, results='hide'}
sf_data <- st_read("MalawiMICS2019-20GPS.shp")

# 2) Read
sf_mics_boundaries <- st_read("mics_boundaries_nr.shp", quiet = TRUE)

#sf_mics_boundaries_2 <- st_read ("1mics_boundaries_nr.shp", queit = TRUE)
```

## 4. Derive Analytic Variables (`wm_clean`)

We recoded all raw MICS variables into analysis-ready categorical
variables, including the composite teenage pregnancy outcome (ever given
birth OR currently pregnant), education level (primary or lower vs.
secondary or higher), wealth index quintile, urbanicity, marital status,
health insurance coverage, contraceptive use, and sexual history. All
factor levels were explicitly defined to ensure consistent ordering in
tables and models.

```{r derive-variables}
wm_clean <- wm_small %>%
  mutate(
    # Ever given birth
    ever_given_birth = case_when(
      cm1 == 1 ~ "Yes",
      cm1 == 2 ~ "No",
      TRUE ~ NA_character_
    ),
    
    # Wealth index quintile
    wealth_index = factor(case_when(
      windex5 == 1 ~ "Poorest",
      windex5 == 2 ~ "Second",
      windex5 == 3 ~ "Middle",
      windex5 == 4 ~ "Fourth",
      windex5 == 5 ~ "Richest"
    ), levels = c("Poorest", "Second", "Middle", "Fourth", "Richest")),
    
    # Current pregnancy
    cur_pregnant = case_when(
      cp1 == 1 ~ "Yes",
      cp1 == 2 ~ "No",
      TRUE ~ NA_character_
    ),
    
    current_marital = case_when(
      ma1 == 1 ~ "Yes - currently married",
      ma1 == 2 ~ "Yes - living with partner",
      ma1 == 3 ~ "No - not in union",
      TRUE     ~ NA_character_
    ),
    
    # Education level (3 categories)
    Edu_level = case_when(
      wb6a %in% c(0) ~ "Primary or lower",
      wb6a %in% c(1) ~ "Primary or lower",
      wb6a %in% c(2:5) ~ "Secondary or higher",
      TRUE ~ NA_character_
    ),
    
    # Urbanicity
    urbanicity = case_when(
      wb16 %in% c(1, 2) ~ "Urban",
      wb16 == 3 ~ "Rural",
      TRUE ~ NA_character_
    ),
    
    # Health insurance
    insurance = case_when(
      wb18 == 1 ~ "Yes",
      wb18 == 2 ~ "No",
      TRUE ~ NA_character_
    ),
    
    # Contraceptive use
    contraceptive_use = case_when(
      cp3 == 1 ~ "Yes",
      cp3 == 2 ~ "No",
      TRUE ~ NA_character_
    ),
    
    # Ever had sex
    had_sex = case_when(
      sb1 >= 8 & sb1 <= 19 ~ "Yes",
      sb1 == 0 ~ "No",
      TRUE ~ NA_character_
    ),
    
    # Teen pregnancy (derived outcome)
    teen_pregnancy = case_when(
      cur_pregnant == "Yes" | ever_given_birth == "Yes" ~ "Yes",
      cur_pregnant == "No" & ever_given_birth == "No" ~ "No",
      TRUE ~ NA_character_
    )
  )
```

## 5. Merge Women's Dataset with Cluster GIS Data

We joined the cleaned women's dataset (`wm_clean`) to the GPS cluster
shapefile on the cluster identifier (`HH1`), creating a
spatially-referenced analytical dataset (`sf_joined`) that retains point
geometry for each respondent's cluster. We verified that coordinate
reference systems were aligned before merging and flagged urbanicity as
a factor variable. The merged object was saved as an RDS file for use in
downstream mapping and spatial regression steps.

```{r merge-women-gis, message=FALSE, warning=FALSE}
# ===============================================================
# Load and Merge MICS Spatial and Survey Data
# ===============================================================

# --- 1. Define main data directory ---
data_dir <- "C:/Users/b.banda/Box/Teenage pregnancy/Lit review Matrix Table/MICS_Analysis/MICS_Analysis"

# --- 2. Read cluster GPS shapefile ---
mics_clusters_sf <- st_read(file.path(data_dir, "MalawiMICS2019-20GPS.shp"), quiet = TRUE)

# --- 3. Read boundary shapefile (district or T/A polygons) ---
mics_boundaries_sf <- st_read(file.path(data_dir, "mics_boundaries/mics_boundaries_nr.shp"), quiet = TRUE)

# --- 4. Check and align coordinate reference systems (CRS) ---
if (st_crs(mics_clusters_sf) != st_crs(mics_boundaries_sf)) {
  mics_clusters_sf <- st_transform(mics_clusters_sf, st_crs(mics_boundaries_sf))
}

# ===============================================================
# 🔹 Merge Women’s Data (wm_clean) with Cluster Spatial Data
# ===============================================================

# ⚠️ Ensure HH1 matches the cluster ID variable in the shapefile
# If your shapefile has a different ID column (e.g., CLUSTERID), adjust the join key
sf_joined <- mics_clusters_sf %>%
  right_join(wm_clean, by = c("HH1" = "hh1")) %>%
  mutate(
    # Create a unique ID for later weighting or identification
    id = sprintf("WM_%04d", row_number()),
    
    # Ensure urbanicity is a proper factor (already derived earlier)
    urbanicity = factor(urbanicity, levels = c("Urban", "Rural"))
  )

# ===============================================================
# 🔹 Quick Data Check
# ===============================================================
sf_joined %>%
  st_drop_geometry() %>%
  count(HH6)

# ===============================================================
# 🔹 Save for later steps (mapping, Moran’s I, regression)
# ===============================================================
saveRDS(sf_joined, file.path(data_dir, "sf_joined_cleaned.RDS"))
```

## 6. Check Missingness

We assessed missing data patterns across all key analysis variables
using both unweighted counts and survey-weighted proportions. We
visualized the missing data pattern using `mice::md.pattern()` and
produced a summary table reporting the number and percentage of missing
values before and after applying sampling weights. This step informed
decisions about whether multiple imputation was warranted for any
variable.

```{r missingness-check, message=FALSE, warning=FALSE}
# ==========================================================
# 🔹 Missingness Analysis for Key Variables
# ==========================================================

# 1. Variables to assess
vars_for_missing <- c(
  "ever_given_birth", "cur_pregnant", "HH6", "insurance",
  "contraceptive_use", "had_sex", "teen_pregnancy", "edu_level", "wmweight"
)

# 2. Subset and standardize
dt_miss <- sf_joined %>%
  select(any_of(vars_for_missing)) %>%
  mutate(
    across(where(is.character) | where(is.logical), as.factor),
    wmweight = suppressWarnings(as.numeric(wmweight))
  )

# 3. Visualize missing data pattern (unweighted)
ok_for_mice <- dt_miss %>%
  select(-wmweight) %>%
  mutate(across(where(is.factor), ~ as.numeric(.x)))

mice::md.pattern(ok_for_mice, rotate.names = TRUE)

# 4. Weighted missingness summary
total_w <- sum(dt_miss$wmweight, na.rm = TRUE)
if (total_w == 0) stop("Survey weights are missing or all zero.")

miss_tbl <- tibble(variable = vars_for_missing) %>%
  mutate(
    n_missing     = map_int(variable, ~ sum(is.na(dt_miss[[.x]]))),
    pct_missing   = round(100 * n_missing / nrow(dt_miss), 1),
    w_missing     = map_dbl(variable, ~ sum(dt_miss$wmweight[is.na(dt_miss[[.x]])], na.rm = TRUE)),
    w_pct_missing = round(100 * w_missing / total_w, 1)
  ) %>%
  arrange(desc(w_pct_missing))

# 5. Display clean summary table
miss_tbl %>%
  gt() %>%
  fmt_number(columns = c(pct_missing, w_pct_missing), decimals = 1) %>%
  tab_header(title = "Weighted and Unweighted Missingness Summary") %>%
  cols_label(
    variable = "Variable",
    n_missing = "N Missing (Unweighted)",
    pct_missing = "% Missing (Unweighted)",
    w_missing = "Weighted N Missing",
    w_pct_missing = "% Missing (Weighted)"
  )
```

## 7. Set Reference Categories and Master Variable Configuration

To ensure consistency across all descriptive tables and regression
models, we define a single master configuration specifying: (i) factor
levels with explicit reference categories ordered first, (ii)
standardised variable labels used across all output tables, and (iii)
the canonical variable order followed in Tables 1–3. Setting reference
categories at the data-preparation stage—rather than at the modelling
stage—guarantees that descriptive tables and odds ratios use identical
orderings, simplifies interpretation, and prevents discrepancies between
the prevalence tables and the multivariable models. Reference categories
were chosen *a priori* based on conceptual considerations (typically the
most-protected or most-prevalent category) to align with prior
literature on adolescent fertility in sub-Saharan Africa.

```{r set-references-and-labels, message=FALSE, warning=FALSE}
# ============================================================
# Master Configuration: Variable Levels, Labels, and Order
# ============================================================
# This block is the single source of truth for variable handling
# throughout the manuscript tables.
# ------------------------------------------------------------

# 1) Reference categories (placed first within each factor)
# ------------------------------------------------------------
# - Edu_level         : "Primary or lower" (modal, lowest education)
# - wealth_index      : "Richest" (highest SES, conceptually protected)
# - HH6               : "Urban" (lower-prevalence residence stratum)
# - current_marital   : "No - not in union" (modal, baseline marital state)
# - insurance         : "No" (most prevalent / unexposed baseline)
# - contraceptive_use : "No" (unexposed baseline)
# - had_sex           : "No" (unexposed baseline)
# ------------------------------------------------------------

sf_joined <- sf_joined %>%
  mutate(
    teen_pregnancy   = factor(teen_pregnancy, levels = c("No", "Yes")),
    Edu_level        = factor(Edu_level,
                              levels = c("Primary or lower", "Secondary or higher")),
    wealth_index     = factor(wealth_index,
                              levels = c("Richest", "Fourth", "Middle", "Second", "Poorest")),
    HH6              = factor(HH6,
                              levels = c("Urban", "Rural")),
    current_marital  = factor(current_marital,
                              levels = c("No - not in union",
                                         "Yes - currently married",
                                         "Yes - living with partner")),
    insurance        = factor(insurance,        levels = c("No", "Yes")),
    contraceptive_use = factor(contraceptive_use, levels = c("No", "Yes")),
    had_sex          = factor(had_sex,          levels = c("No", "Yes"))
  )

# 2) Master label map (single source of truth across all tables)
# ------------------------------------------------------------
master_labels <- c(
  Edu_level         = "Education",
  wealth_index      = "Wealth Index",
  HH6               = "Place of Residence",
  current_marital   = "Marital status",
  contraceptive_use = "Contraceptive use",
  insurance         = "Health insurance",
  had_sex           = "Ever had sex"
)

# 3) Canonical variable order across Tables 1, 2, 3
# ------------------------------------------------------------
master_order <- c(
  "Edu_level",
  "wealth_index",
  "HH6",
  "current_marital",
  "contraceptive_use",
  "insurance"
)

# 4) Master factor-level lookup (kept here for downstream label-aware loops)
# ------------------------------------------------------------
master_levels <- list(
  Edu_level         = c("Primary or lower", "Secondary or higher"),
  wealth_index      = c("Richest", "Fourth", "Middle", "Second", "Poorest"),
  HH6               = c("Urban", "Rural"),
  current_marital   = c("No - not in union", "Yes - currently married", "Yes - living with partner"),
  contraceptive_use = c("No", "Yes"),
  insurance         = c("No", "Yes"),
  had_sex           = c("No", "Yes")
)
```

## 8. Create Survey Design Object

We specified the complex survey design using `svydesign()`, defining
enumeration area clusters (`HH1`) as the primary sampling unit,
district-by-urbanicity combinations as the stratification variable
(`stratum`), and individual woman sampling weights (`wmweight`). Setting
`nest = TRUE` accounts for cluster identifiers being nested within
strata. This design object (`mics_svy`) underpins all survey-weighted
prevalence estimates, chi-square tests, and logistic regression models
in Parts III and IV. The design carries the factor-level orderings
defined above, ensuring reference categories appear first in every
downstream table.

```{r survey-design}
mics_svy <- svydesign(
  id      = ~HH1,
  strata  = ~stratum,
  weights = ~wmweight,
  data    = sf_joined,
  nest    = TRUE
)

# Quick verification: confirm reference levels are first
cat("Verification of reference categories (first level of each factor):\n")
purrr::walk(master_order, function(v) {
  lvls <- levels(mics_svy$variables[[v]])
  cat(sprintf("  %-20s reference = %s\n", v, lvls[1]))
})
```

## 9. Helper Functions for Survey-weighted Tables

We defined a set of reusable helper functions to streamline
survey-weighted estimation throughout the analysis. `svyciprop_fun()`
computes a survey-weighted proportion with likelihood-based 95%
confidence intervals for a single variable level. `svyciprop_by_fun()`
extends this to stratified estimates. `style_pvalue2()` formats p-values
to publication standard. `model_log()`, `model_lin()`, and
`model_pois()` are wrappers around `svyglm()` for logistic, linear, and
Poisson regression under the complex survey design.
`add_significance_stars()` is used downstream to format `gtsummary`
regression tables with conventional significance markers (*p\<0.05;
**p\<0.01;*** p\<0.001).

```{r help-functions}
# Thousands-formatting helper used in all table footers and headers
comma <- function(x) format(x, big.mark = ",", scientific = FALSE)

# Significance-stars helper for gtsummary regression tables
add_significance_stars <- function(tbl,
                                   pattern = "{estimate} [{conf.low}-{conf.high}]{stars}",
                                   hide_p  = TRUE,
                                   hide_se = TRUE) {
  tbl <- tbl %>%
    modify_table_body(
      ~ .x %>%
        mutate(
          stars = case_when(
            !is.na(p.value) & p.value < 0.001 ~ "***",
            !is.na(p.value) & p.value < 0.01  ~ "**",
            !is.na(p.value) & p.value < 0.05  ~ "*",
            TRUE                              ~ ""
          )
        )
    ) %>%
    modify_column_merge(pattern = pattern, rows = !is.na(estimate))

  if (hide_p  && "p.value"   %in% names(tbl$table_body)) tbl <- tbl %>% modify_column_hide(columns = p.value)
  if (hide_se && "std.error" %in% names(tbl$table_body)) tbl <- tbl %>% modify_column_hide(columns = std.error)

  tbl
}

svyciprop_fun <- function(design, var, var_lvl) {
  # Build formula for indicator: e.g., I(gender == "Man")
  form <- paste0("I(", var, " == ", deparse(var_lvl), ")")
  
  # Compute the weighted proportion using svyciprop() with likelihood method
  svy_props <- svyciprop(
    as.formula(paste0("~", form)),
    design,
    method = "likelihood"
  )
  
  # Extract the proportion and confidence intervals, then scale to percentages
  prop <- svy_props[1]
  cil <- attr(svy_props, "ci")[1]
  ciu <- attr(svy_props, "ci")[2]
  
  df_out <- data.frame(
    var = var,
    lvl = var_lvl,
    est = prop,
    ci_l = cil,
    ci_u = ciu
  ) %>%
    mutate(
      est = est * 100,
      ci_l = ci_l * 100,
      ci_u = ci_u * 100
    )
  
  return(df_out)
}

svyciprop_by_fun <- function(.data, var, var_lvl, by) {
  form <- paste0("I(", var, " == ", deparse(var_lvl), ")")
  
  svy_props <- svyby(
    as.formula(paste0("~", form)),
    by = as.formula(paste0("~", by)),
    design = .data,
    svyciprop,
    vartype = "ci",
    method = "likelihood"
  ) %>%
    rename("est" = form) %>%
    mutate(
      var = var,
      lvl = var_lvl,
      lvl = factor(lvl, levels = var_lvl),
      est = est * 100,
      ci_l = ci_l * 100,
      ci_u = ci_u * 100
    )
  
  return(svy_props)
}

miss_svy <- function(.data, var, dkrf = TRUE) {
  na_len <- sum(is.na(.data[, var]))
  var_formal_lc <- var
  
  if (dkrf == TRUE) {
    .var <- sym(var)
    .data_sub <- filter(.data, !!.var %in% c("Don't know/Refused", "Not applicable"))
    dkrf_len <- sum(!is.na(.data_sub[, .var]))
    na_len <- na_len + dkrf_len
  }
  
  statement <- paste0("Participants missing ", var_formal_lc, " were removed (N = ", na_len, ").")
  return(statement)
}

style_pvalue2 <- function(x, digits = 2, prepend_p = FALSE,
                          big.mark = NULL, decimal.mark = ".", ...) {

  # Pre-allocate result vector
  p_fmt <- character(length(x))

  # Loop through each value in x
  for (i in seq_along(x)) {
    xi <- x[i]

    if (is.na(xi)) {
      p_fmt[i] <- NA_character_
    } else if (digits == 1) {
      p_fmt[i] <- dplyr::case_when(
        xi > 0.9 ~ paste0(">", format(0.9, digits = 1, nsmall = 1)),
        round(xi, 1) >= 0.2 ~ format(xi, digits = 1, nsmall = 1),
        round(xi, 2) >= 0.1 ~ format(xi, digits = 2, nsmall = 2),
        xi >= 0.001 ~ format(xi, digits = 3, nsmall = 3),
        xi < 0.001 ~ "<0.001"
      )
    } else if (digits == 2) {
      p_fmt[i] <- dplyr::case_when(
        xi > 0.99 ~ ">0.99",
        round(xi, 2) >= 0.1 ~ format(xi, digits = 2, nsmall = 2),
        xi >= 0.01 ~ format(xi, digits = 2, nsmall = 2),
        xi >= 0.001 ~ "<0.01",
        xi < 0.001 ~ "<0.001"
      )
    } else if (digits == 3) {
      p_fmt[i] <- dplyr::case_when(
        xi > 0.999 ~ ">0.999",
        xi >= 0.001 ~ format(xi, digits = 3, nsmall = 3),
        xi < 0.001 ~ "<0.001"
      )
    } else if (digits == 4) {
      p_fmt[i] <- dplyr::case_when(
        xi > 0.9999 ~ ">0.9999",
        xi >= 0.0001 ~ format(xi, digits = 4, nsmall = 4),
        xi < 0.0001 ~ "<0.0001"
      )
    } else {
      stop("The `digits=` argument must be 1, 2, 3, or 4.")
    }

    if (prepend_p && !is.na(p_fmt[i])) {
      if (startsWith(p_fmt[i], "<") || startsWith(p_fmt[i], ">")) {
        p_fmt[i] <- paste0("p", p_fmt[i])
      } else {
        p_fmt[i] <- paste0("p=", p_fmt[i])
      }
    }
  }

  return(p_fmt)
}

# Survey-weighted unadjusted logistic regression.
# Ensures correct standard errors under complex survey design.
model1 <- function(.data, outcome, ...) {
  .form <- as.formula(paste0(outcome, " ~ ", paste(..., collapse = "+")))
  svyglm(.form, design = .data, family = "quasibinomial")
}

# Survey-weighted linear regression for continuous outcomes.
# Accounts for weights, strata, and clustering in estimating means.
model_lin <- function(.data, outcome, ...) {
  .form <- as.formula(paste0(outcome, " ~ ", paste(..., collapse = "+")))
  svyglm(.form, design = .data, family = "gaussian")
}

# Survey-weighted logistic regression with multiple predictors.
# Returns robust estimates using the quasibinomial family.
model_log <- function(.data, outcome, ...) {
  .form <- as.formula(paste0(outcome, " ~ ", paste(..., collapse = "+")))
  svyglm(.form, design = .data, family = "quasibinomial")
}

# Poisson regression with survey weights for count data outcomes.
# Recommended for modeling counts like symptom scores or events.
model_pois <- function(.data, outcome, ...) {
  .form <- as.formula(paste0(outcome, " ~ ", paste(..., collapse = "+")))
  svyglm(.form, design = .data, family = "quasipoisson")
}
```

# PART III: Risk Factor Analysis - Final Tables

The three tables below are the publication-ready tables for the paper.
All estimates are survey-weighted to account for the complex MICS
sampling design. Sample sizes (n) are weighted and rounded to the
nearest whole number, representing the population these respondents
represent. All percentages, odds ratios, and confidence intervals are
derived from survey-weighted analyses.

## Final Table 1: Sample Demographics

We constructed a sample-characteristics table presenting weighted sample
sizes (rounded to the nearest whole number) and survey-weighted column
percentages with 95% confidence intervals for all sociodemographic and
behavioural exposures, stratified by teenage pregnancy status (Yes/No)
alongside the overall analytic sample. Categories within each variable
are ordered with the reference category (used in the regression in Table
3) appearing first; this ensures direct visual alignment between Tables
1 and 3. Sample sizes vary slightly across variables due to item-level
non-response; missingness for each variable is reported in Section 6
(Missingness Analysis).

```{r table1-complete, message=FALSE, warning=FALSE}
# ============================================================
# Table 1: Sample Demographics by Teen Pregnancy Status
# ============================================================
# Uses master_order, master_labels, master_levels defined in Section 7.
# `comma()` and `add_significance_stars()` are defined in Section 9.
# Complete  Table 1 stratified by teen pregnancy


comma <- function(x) format(x, big.mark = ",", scientific = FALSE)

# # --- Variable levels and labels (excluding teen_pregnancy) ------------------
tls <- list(

  Edu_level         = c("Primary or lower","Secondary or higher"),
  HH6               = c("Rural", "Urban"),
  insurance         = c("No", "Yes"),
  contraceptive_use = c("No", "Yes"),
  wealth_index      = c("Poorest", "Second", "Middle", "Fourth", "Richest"),
  current_marital   = c("Yes - currently married", "Yes - living with partner", "No - not in union"))

# Define variable label mapping
var_labels <- c(
  Edu_level          = "Education Level",
  HH6                = "Urbanicity",
  wealth_index       = "Wealth Index",
  current_marital    = "Current Marital Status",
   contraceptive_use  = "Contraceptive Use",
   insurance          = "Health Insurance Coverage",
  ever_given_birth   = "Ever Given Birth")

# Define total_n used in cols_label()
total_n <- nrow(mics_svy$variables)

# --- Overall (Total) column with unweighted counts --------------------------
overall_tbl <- map_dfr(
  names(var_labels),
  function(v) {
    lvls <- tls[[v]]
    if (length(lvls) == 0) {
      message("Warning: No valid levels for ", v)
      return(tibble(var = v, lvl = character(), est_tot = character(), n_tot = integer()))
    }
    prop <- map_dfr(lvls, ~ {
      res <- tryCatch(
        svyciprop_fun(mics_svy, v, .x),
        error = function(e) {
          message("Warning: svyciprop_fun failed for ", v, "=", .x, ": ", e$message)
          return(tibble(est = NA, ci_l = NA, ci_u = NA))
        }
      )
      res %>% mutate(lvl = .x)
    })
    count <- map_dfr(lvls, ~ {
      dat <- mics_svy$variables
      n <- sum(dat[[v]] == .x, na.rm = TRUE)
      tibble(lvl = .x, n_tot = n)
    })
    prop %>%
      left_join(count, by = "lvl") %>%
      mutate(var = v, lvl = factor(lvl, levels = lvls)) %>%
      arrange(lvl) %>%
      mutate(
        est_tot = case_when(
          is.na(est) | is.na(ci_l) | is.na(ci_u) ~ sprintf("NA (%s) [NA, NA]", comma(n_tot)),
          TRUE ~ sprintf("%0.1f%% (%s) [%0.1f%%, %0.1f%%]", est, comma(n_tot),  ci_l, ci_u)
        )
      ) %>%
      select(var, lvl, est_tot, n_tot)
  }
)

# --- Stratified by teen_pregnancy (No/Yes) with unweighted counts -----------
by_tbl <- map_dfr(
  names(var_labels),
  function(v) {
    lvls <- tls[[v]]
    if (length(lvls) == 0) {
      message("Warning: No valid levels for ", v)
      return(tibble(var = v, lvl = character(), No = character(), Yes = character(), n_no = integer(), n_yes = integer()))
    }
    prop <- map_dfr(lvls, ~ {
      lv <- .x
      fstr <- paste0("I(", v, " == \"", lv, "\")")
      out <- tryCatch(
        {
          res <- svyby(
            as.formula(paste0("~", fstr)),
            ~teen_pregnancy,
            mics_svy, 
            svyciprop, 
            method = "logit",
            vartype = "ci", 
            na.rm = TRUE, 
            keep.names = TRUE
          )
          names(res)[names(res) == fstr] <- "est"
          names(res)[names(res) == paste0(fstr, ".lower")] <- "ci_l"
          names(res)[names(res) == paste0(fstr, ".upper")] <- "ci_u"
          res
        },
        error = function(e) {
          message("Warning: svyby failed for ", v, "=", lv, " by teen_pregnancy: ", e$message)
          return(data.frame(teen_pregnancy = c("No", "Yes"), 
                            est = NA, ci_l = NA, ci_u = NA))
        }
      )
      out %>%
        mutate(lvl = lv) %>%
        select(teen_pregnancy, lvl, est, ci_l, ci_u)
    })
    count <- map_dfr(lvls, ~ {
      dat <- mics_svy$variables
      n <- tapply(dat[[v]] == .x, dat[["teen_pregnancy"]], sum, na.rm = TRUE)
      data.frame(
        teen_pregnancy = names(n),
        lvl = .x,
        n = as.numeric(n)
      ) %>%
        filter(!is.na(teen_pregnancy))
    })
    prop %>%
      left_join(count, by = c("teen_pregnancy", "lvl")) %>%
      mutate(
        var = v,
        est_ci = case_when(
          is.na(est) | is.na(ci_l) | is.na(ci_u) ~ sprintf("NA (%s) [NA, NA]", comma(n)),
          TRUE ~ sprintf("%0.1f%% (%s) [%0.1f%%, %0.1f%%]", 100 * est, comma(n), 100 * ci_l, 100 * ci_u)
        )
      ) %>%
      select(var, lvl, teen_pregnancy, est_ci, n) %>%
      pivot_wider(names_from = teen_pregnancy, values_from = c(est_ci, n), values_fill = list(est_ci = "", n = 0)) %>%
      mutate(lvl = factor(lvl, levels = lvls)) %>%
      arrange(lvl) %>%
      rename(n_no = n_No, n_yes = n_Yes)
  }
)

# --- Combine & render table -------------------------------------------------
final_tbl <- overall_tbl %>%
  left_join(by_tbl, by = c("var", "lvl")) %>%
  mutate(var_label = var_labels[var],
         lvl = as.character(lvl)) %>%
  select(var_label, lvl, n_yes, Yes = est_ci_Yes,n_no, No = est_ci_No,n_tot, est_tot)

gt_tbl1_strat <- final_tbl %>%
  gt(groupname_col = "var_label") %>%
  tab_header(
    title = "Table 1: Sample demographics",
    subtitle = "Unweighted total N and survey-weighted percentages (95% CI), stratified by Teen Pregnancy"
  ) %>%
  cols_label(
    lvl = "Level",
    n_tot = "n",
    est_tot = html(paste0("Total (Unweighted n = ", comma(total_n), ")")),
    n_no = "n = 4,175",
    No = "Teen Pregnancy: No",
    n_yes = "n = 1,595",
    Yes = "Teen Pregnancy: Yes"
  ) %>%
  tab_footnote(
    footnote = "Estimates and confidence intervals are survey-weighted; 95% CIs use Rao–Scott adjustments where applicable.",
    locations = cells_title(groups = "title")
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.align = "left",
    column_labels.font.weight = "bold"
  )

# Display the table
gt_tbl1_strat
```

## Supplementary Table S1: Survey-Weighted Chi-Square Tests

We tested the bivariate association between teenage pregnancy and each
covariate using chi-square tests on survey-weighted contingency tables
(`svytable()` + `chisq.test()`). This approach correctly propagates
sampling weights into the cell counts without the matrix decomposition
issues that can arise with Rao-Scott corrections on sparse multi-level
variables. P-values below 0.05 are bolded to highlight statistically
significant associations. Variables and labels follow the master
configuration in Section 7.

```{r table-s1-chisq, message=FALSE, warning=FALSE}
# ============================================================
# Supplementary Table S1: Survey-Weighted Chi-Square Tests
# ============================================================

chi_results <- map_dfr(master_order, function(v) {
  sub_design <- subset(mics_svy, !is.na(teen_pregnancy) & !is.na(get(v)))

  tab  <- svytable(as.formula(paste0("~teen_pregnancy + ", v)), design = sub_design)
  test <- chisq.test(tab)

  tibble(
    Variable = master_labels[v],
    Chi_sq   = as.numeric(test$statistic),
    df       = as.numeric(test$parameter),
    p_value  = test$p.value
  )
})

chi_results %>%
  gt() %>%
  tab_header(
    title    = md("**Supplementary Table S1.** Survey-weighted χ² tests of teenage pregnancy vs sociodemographic and behavioural covariates"),
    subtitle = md("Malawi MICS 2019–20")
  ) %>%
  fmt_number(columns = Chi_sq,  decimals = 2) %>%
  fmt_number(columns = df,      decimals = 0) %>%
  fmt(columns = p_value, fns = function(x) style_pvalue2(x, digits = 3)) %>%
  cols_label(
    Variable = "Covariate",
    Chi_sq   = md("χ² statistic"),
    df       = "df",
    p_value  = "p-value"
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_body(columns = p_value, rows = p_value < 0.05)
  ) %>%
  tab_footnote(
    footnote  = paste(
      "Tests use Pearson χ² applied to survey-weighted contingency tables (svytable), which propagates sampling weights into cell counts under the complex MICS design.",
      "Bold p-values indicate significance at α = 0.05.",
      "Each test is restricted to respondents with non-missing values on both teenage pregnancy and the covariate."
    ),
    locations = cells_title(groups = "title")
  ) %>%
  tab_options(
    table.font.size           = px(13),
    heading.align             = "left",
    column_labels.font.weight = "bold"
  )
```

## Final Table 2: Weighted Prevalence of Teenage Pregnancy

We estimated the survey-weighted prevalence of teenage pregnancy (%, 95%
CI) within each level of every exposure variable using `svyby()` with
`svyciprop` (logit-transformed CIs). Each row reports the weighted count
of pregnant teens, the total weighted sample within that stratum, and
the within-stratum prevalence with its 95% CI. Variables are presented
in the same canonical order as Tables 1 and 3, and reference categories
appear first to facilitate cross-table comparison.

```{r table2-prevalence, message=FALSE, warning=FALSE}
# --- Variables for prevalence table
target_vars <- c(
  "Edu_level", "HH6","wealth_index",  "current_marital", "insurance", 
 "contraceptive_use","had_sex"

)

var_labels <- c(
  ever_given_birth   = "Ever Given Birth",
  GEONAMES           = "District",
  cur_pregnant       = "Currently Pregnant",
  Edu_level          = "Education Level",
  HH6                = "Place of residence",
  insurance          = "Health Insurance Coverage",
  contraceptive_use  = "Contraceptive Use",
  wealth_index       = "Wealth Index",
  current_marital    = "Current Marital Status",
  had_sex            =  "Ever had sex"
)

# --- 1) Weighted prevalence of teen pregnancy (teen_pregnancy == "Yes")
prev_tbl <- map_dfr(target_vars, function(v) {
  dat <- mics_svy$variables
  lvls <- unique(na.omit(as.character(dat[[v]])))
  if (length(lvls) == 0) return(NULL)
  
  svy_res <- svyby(
    ~I(teen_pregnancy == "Yes"),
    as.formula(paste0("~", v)),
    mics_svy,
    svyciprop,
    vartype = "ci",
    method = "logit",
    na.rm = TRUE
  )
  
  est_col <- setdiff(names(svy_res), c(v, "ci_l", "ci_u", "se", "SE"))
  if (length(est_col) == 0) stop("No estimate column found.")
  
  svy_res %>%
    rename(est = all_of(est_col[1])) %>%
    mutate(
      var = v,
      lvl = as.character(!!sym(v)),
      est_ci = sprintf("%0.1f%% [%0.1f%%, %0.1f%%]",
                       100 * est, 100 * ci_l, 100 * ci_u)
    ) %>%
    select(var, lvl, est, ci_l, ci_u, est_ci)
}) %>%
  mutate(lvl = as.character(lvl))

# --- 2) Unweighted total counts for each level
unweighted_counts <- map_dfr(target_vars, function(v) {
  mics_svy$variables %>%
    filter(!is.na(.data[[v]])) %>%
    mutate(lvl = as.character(.data[[v]])) %>%
    count(lvl, name = "n_unweighted") %>%
    mutate(var = v)
})

# --- 3) NEW: Unweighted counts of teen pregnancy = "Yes"
teen_preg_counts <- map_dfr(target_vars, function(v) {
  mics_svy$variables %>%
    filter(!is.na(.data[[v]])) %>%
    mutate(lvl = as.character(.data[[v]])) %>%
    group_by(lvl) %>%
    summarise(
      n_teen_preg = sum(teen_pregnancy == "Yes", na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(var = v)
})

# --- 4) Merge counts + labels
prev_tbl_final <- prev_tbl %>%
  left_join(unweighted_counts, by = c("var", "lvl")) %>%
  left_join(teen_preg_counts, by = c("var", "lvl")) %>%
  mutate(var_label = var_labels[var]) %>%
  select(var_label, lvl, n_unweighted, n_teen_preg, est_ci)

# --- 5) Render as GT table
gt_prev_teen_pregnancy <- prev_tbl_final %>%
  gt(groupname_col = "var_label") %>%
  tab_header(
    title = "Table 2. Prevalence of Teenage Pregnancy by Sociodemographic Characteristics",
    subtitle = "Unweighted total N, unadjusted teen pregnancy counts, and survey-weighted % (95% CI)"
  ) %>%
  cols_label(
    lvl = "Level",
    n_unweighted = html(paste0("Total Participants (n = ", comma(nrow(mics_svy$variables)), ")")),
    n_teen_preg = "Teenage pregnancies (n= 1595)",
    est_ci = "Weighted % [95% CI]"
  ) %>%
  tab_footnote(
    footnote = paste(   "Source: Malawi Multiple Indicator Cluster Survey (MICS) 2019–20, women's questionnaire.",
      "Teen pregnancy defined as ever having given birth (CM1 = 1) or being currently pregnant (CP1 = 1).",
      " Estimates and confidence intervals are survey-weighted; 95% CIs use Rao–Scott adjustments, accounting for the two-stage stratified cluster design."
    )
,
    locations = cells_title(groups = "title")
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.align = "left",
    column_labels.font.weight = "bold"
  )

gt_prev_teen_pregnancy
```

## Final Table 3: Survey-weighted Logistic Regression

We estimated associations between each exposure and teenage pregnancy
using survey-weighted logistic regression (`svyglm`, quasibinomial
family) under two specifications: Model 1 reports unadjusted odds ratios
from separate univariable models for each covariate; Model 2 reports
adjusted odds ratios from a single multivariable model including all
covariates. Reference categories follow the master configuration in
Section 7. The multivariable model uses complete cases on all included
covariates; the analytic sample size is reported in the table footer.

```{r table3-regression, message=FALSE, warning=FALSE}
# ============================================================
# Table 3: Logistic Regression of Teenage Pregnancy
# Among Adolescent Females (15–19 Years), Malawi MICS 2019–20
# ============================================================
# Reference categories and variable order are inherited from
# the master configuration in Section 7. Only the binary
# numeric outcome is created here.
# ------------------------------------------------------------

# Update design to add binary numeric outcome
mics_svy <- update(mics_svy, teen_pregnancy_num = as.numeric(teen_pregnancy == "Yes"))

# Helper: gtsummary label list (formula form) for adjusted model
full_label_list <- map2(master_order, master_labels[master_order],
                        ~ as.formula(paste0(.x, " ~ '", .y, "'")))

# Helper: single-variable label for univariable models
make_single_label <- function(var) {
  stats::setNames(list(master_labels[[var]]), var)
}

# ------------------------------------------------------------
# Step 1: Univariable (Model 1) survey-weighted logistic models
# ------------------------------------------------------------
unadj_models <- map(master_order, function(v) {
  svyglm(
    as.formula(paste0("teen_pregnancy_num ~ ", v)),
    design = mics_svy,
    family = quasibinomial()
  )
}) %>% set_names(master_order)

# ------------------------------------------------------------
# Step 2: Multivariable (Model 2) adjusted survey-weighted model
# Restricted internally by svyglm to complete cases on all
# covariates plus the outcome.
# ------------------------------------------------------------
adj_formula <- as.formula(
  paste0("teen_pregnancy_num ~ ", paste(master_order, collapse = " + "))
)
adj_model_full <- svyglm(adj_formula, design = mics_svy, family = quasibinomial())

# Analytic sample size of the adjusted model (complete cases)
adj_n_unweighted <- length(adj_model_full$residuals)
adj_n_weighted   <- ceiling(sum(mics_svy$variables$wmweight[
  rowSums(is.na(mics_svy$variables[, c("teen_pregnancy_num", master_order)])) == 0
], na.rm = TRUE))

# ------------------------------------------------------------
# Step 3: Format Model 1 (unadjusted ORs)
# ------------------------------------------------------------
unadj_tp_log <- map(
  master_order,
  function(j) {
    unadj_models[[j]] %>%
      tbl_regression(
        exponentiate = TRUE,
        intercept    = FALSE,
        pvalue_fun   = ~ style_pvalue2(.x, digits = 2),
        label        = make_single_label(j)
      ) %>%
      add_significance_stars(
        pattern = "{estimate} [{conf.low}-{conf.high}]{stars}",
        hide_p  = TRUE,
        hide_se = TRUE
      )
  }
) %>%
  tbl_stack()

# ------------------------------------------------------------
# Step 4: Format Model 2 (adjusted aORs)
# ------------------------------------------------------------
adj_tp_log_full <- adj_model_full %>%
  tbl_regression(
    exponentiate = TRUE,
    intercept    = FALSE,
    pvalue_fun   = ~ style_pvalue2(.x, digits = 2),
    label        = full_label_list
  ) %>%
  add_significance_stars(
    pattern = "{estimate} [{conf.low}-{conf.high}]{stars}",
    hide_p  = TRUE,
    hide_se = TRUE
  )

# ------------------------------------------------------------
# Step 5: Merge models into one publication table
# ------------------------------------------------------------
teen_preg_log <- tbl_merge(
  tbls = list(unadj_tp_log, adj_tp_log_full),
  tab_spanner = c(
    "**Model 1: Unadjusted OR [95% CI]**",
    "**Model 2: Adjusted aOR [95% CI]**"
  )
) %>%
  modify_header(
    label      ~ "**Characteristic**",
    estimate_1 ~ "**Model 1: Unadjusted OR [95% CI]**",
    estimate_2 ~ "**Model 2: Adjusted aOR [95% CI]**"
  ) %>%
  modify_caption(
    "**Table 3.** Survey-weighted logistic regression of teenage pregnancy among adolescent females (15–19 years), Malawi MICS 2019–20"
  ) %>%
  bold_labels() %>%
  modify_spanning_header(
    everything() ~ "**Teenage Pregnancy**"
  ) %>%
  modify_footnote(
    everything() ~ paste0(
      "Significance: *p<0.05; **p<0.01; ***p<0.001. ",
      "Estimates derived from survey-weighted logistic regression (quasi-binomial) accounting for the two-stage stratified cluster design ",
      "(PSU = HH1, strata = district × urbanicity, weight = wmweight). ",
      "Model 1 presents unadjusted odds ratios from separate univariable models for each covariate. ",
      "Model 2 presents adjusted odds ratios from a single multivariable model including all listed covariates ",
      "(complete-case analytic sample, unweighted n = ", comma(adj_n_unweighted),
      "; weighted n = ", comma(adj_n_weighted), "). ",
            "Teen pregnancy defined as ever given birth (CM1 = 1) or currently pregnant (CP1 = 1). ",
      "OR = odds ratio; aOR = adjusted odds ratio; CI = confidence interval."
    )
  )

teen_preg_log
```

# PART IV: Spatial Analysis - Hotspots of Teenage Pregnancy

We prepared the spatial analysis dataset by creating a respondent-level
analytical object (`teens`) from `wm_clean`, filtered to women with
non-missing pregnancy status. We linked each respondent to her
Traditional Authority name (`ta_name`) by joining on the cluster
identifier (`HH1`) to the GPS cluster shapefile, which contains the
`GEONAMET` field assigned during data collection.

```{r spatial-prep-libs, message=FALSE, warning=FALSE}
# Load districts shapefile for overlay on LISA map
districts <- st_read("mics_boundaries3.shp", quiet = TRUE)

# Join with sf_data on HH1 to attach ta_name (GEONAMET) from the GPS cluster file
teens <- wm_clean %>%
  rename_with(tolower) %>%
  filter(!is.na(teen_pregnancy)) %>%
  mutate(pregnant = if_else(teen_pregnancy == "Yes", 1L, 0L)) %>%
  left_join(
    sf_data %>% st_drop_geometry() %>% select(HH1, ta_name = GEONAMET),
    by = c("hh1" = "HH1")
  )
```

## 0. Link clusters to Traditional Authorities

We loaded the TA boundary shapefile and linked each survey cluster in
`sf_joined` to its Traditional Authority using a table join on `HH1`
rather than a spatial join, since the GPS cluster file already contains
the `GEONAMET` TA assignment. We constructed a separate survey design
object (`mics_svy_ta`) that includes `ta_name`, preserving the original
`mics_svy` object for the risk factor tables in Part III.

```{r link-clusters-to-ta, message=FALSE, warning=FALSE}
cat("Loading TA boundaries and linking clusters to TAs...\n")

ta_boundaries <- st_read("mics_boundaries_nr.shp", quiet = TRUE) %>%
  select(ta_name = GEONAMET, geometry) %>%
  mutate(ta_name = trimws(ta_name)) %>%
  filter(ta_name != "") %>%
  st_make_valid()

cat("TA boundaries loaded:", nrow(ta_boundaries), "records\n")

# Table join: link each cluster to its TA using HH1 -> GEONAMET from GPS file
sf_joined_with_ta <- sf_joined %>%
  left_join(
    sf_data %>% st_drop_geometry() %>% select(HH1, ta_name = GEONAMET),
    by = "HH1"
  ) %>%
  mutate(ta_name = replace_na(ta_name, "Unknown"))

cat("Clusters linked to TAs\n")
cat("  N clusters in Unknown TA:", sum(sf_joined_with_ta$ta_name == "Unknown"), "\n")

# Separate survey design for spatial analysis (preserves original mics_svy for Tables 1-3)
mics_svy_ta <- svydesign(
  id      = ~HH1,
  strata  = ~stratum,
  weights = ~wmweight,
  data    = sf_joined_with_ta,
  nest    = TRUE
)

cat("Survey design mics_svy_ta built with ta_name variable\n")
```

## 1. Calculate TA-Level Survey-Weighted Prevalence

We used `svyby()` with `svymean` to estimate the survey-weighted
prevalence of teenage pregnancy within each Traditional Authority, using
`mics_svy_ta` which carries the `ta_name` grouping variable. We
extracted point estimates and 95% confidence intervals and converted all
proportions to percentages for reporting.

```{r ta-prevalence, message=FALSE, warning=FALSE}
cat("\nCalculating survey-weighted prevalence by TA...\n")

ta_prev_svy <- svyby(
  ~teen_pregnancy,
  ~ta_name,
  mics_svy_ta,
  svymean,
  vartype = "ci",
  na.rm = TRUE
)

names(ta_prev_svy)

ta_prev <- ta_prev_svy %>%
  as_tibble() %>%
  rename(
    prev = teen_pregnancyYes,
    ci_low = ci_l.teen_pregnancyYes,
    ci_high = ci_u.teen_pregnancyYes
  ) %>%
  mutate(
    prev_pct = prev * 100,
    ci_low_pct = ci_low * 100,
    ci_high_pct = ci_high * 100
  )
```

## 2. Add TA-Level Sample Sizes

We computed unweighted sample sizes per Traditional Authority from the
`teens` object, counting the total number of adolescent respondents and
the number with a positive pregnancy outcome. These counts were merged
with the weighted prevalence estimates to allow assessment of estimate
precision and to flag TAs with small samples (fewer than 10 or 20
respondents).

```{r ta-sample-sizes, message=FALSE, warning=FALSE}
ta_n <- teens %>%
  group_by(ta_name) %>%
  summarise(
    n_teen = n(),
    n_pregnant = sum(pregnant, na.rm = TRUE),
    .groups = "drop"
  )

ta_data <- ta_prev %>%
  left_join(ta_n, by = "ta_name") %>%
  arrange(desc(prev_pct)) %>%
  mutate(
    flag_small = n_teen < 20,
    flag_very_small = n_teen < 10
  )

cat("\nTA-LEVEL PREVALENCE: Top 10 by prevalence\n")
print(ta_data %>% head(10))
```

## 3. Sample Size Quality Check

We printed a summary of TA-level sample adequacy to identify units with
very small samples that may produce unstable prevalence estimates. All
TAs with at least one respondent were retained in the spatial analysis
to avoid selection bias, but small-sample TAs are flagged in the output
table.

```{r sample-quality-check, message=FALSE, warning=FALSE}
cat("\n", strrep("=", 80), "\n", sep = "")
cat("SAMPLE SIZE QUALITY CHECK\n")
cat(strrep("=", 80), "\n", sep = "")
cat("TAs with <10 teens, very small:", sum(ta_data$flag_very_small), "\n")
cat("TAs with <20 teens, small:", sum(ta_data$flag_small), "\n")
cat("Mean n per TA:", round(mean(ta_data$n_teen), 1), "\n")
cat("Median n per TA:", median(ta_data$n_teen), "\n")
```

## 4. Clean and Aggregate Boundary Polygons

We excluded TA Nkula due to topology errors and dissolved multi-part
polygons for TAs that appear as more than one feature in the boundary
shapefile using `st_combine()`. This ensures a one-to-one match between
the prevalence data and boundary features when merging for spatial
analysis.

```{r clean-boundaries, message=FALSE, warning=FALSE}
ta_boundaries_clean <- ta_boundaries %>%
  filter(ta_name != "TA Nkula")

ta_boundaries_unique <- ta_boundaries_clean %>%
  group_by(ta_name) %>%
  summarise(geometry = st_combine(geometry), .groups = "drop") %>%
  st_make_valid()
```

## 5. Merge Prevalence Data with Boundary Polygons

We joined the TA-level weighted prevalence estimates to the cleaned
boundary polygons to create the primary spatial analysis object
(`ta_lisa`). We then computed a Z-score standardization of prevalence
(`prev_z`) required for the Moran scatter plot and LISA quadrant
classification. TAs without prevalence data were excluded at this step.

```{r merge-prev-boundaries, message=FALSE, warning=FALSE}
ta_lisa <- ta_boundaries_unique %>%
  left_join(ta_data, by = "ta_name") %>%
  filter(!is.na(prev)) %>%
  mutate(
    prev_z = (prev - mean(prev, na.rm = TRUE)) / sd(prev, na.rm = TRUE),
    prev_z = replace_na(prev_z, 0)
  )
```

## 6. Merge Summary Check

We printed a record-level summary verifying the number of TA boundary
units at each stage of the cleaning and merging pipeline to confirm that
no records were unexpectedly dropped or duplicated before spatial
analysis.

```{r merge-summary, message=FALSE, warning=FALSE}
cat("\nMerged dataset summary:\n")
cat("Boundary units after excluding TA Nkula:", nrow(ta_boundaries_clean), "\n")
cat("Unique boundary units after combining duplicates:", nrow(ta_boundaries_unique), "\n")
cat("Units with prevalence data:", n_distinct(ta_data$ta_name), "\n")
cat("Units included in spatial analysis:", nrow(ta_lisa), "\n")
```

## 7. Create KNN-5 Spatial Weights

We defined spatial relationships between Traditional Authorities using
K-nearest neighbours (K = 5) based on TA polygon centroids. Weights were
row-standardised (style = `"W"`) so that each TA's neighbours contribute
equally regardless of how many neighbours it has. This weight matrix is
used in both the global Moran's I test and the LISA analysis.

```{r knn5-weights, message=FALSE, warning=FALSE}
ta_centers <- st_centroid(ta_lisa)
coords <- st_coordinates(ta_centers)
knn5 <- knearneigh(coords, k = 5)
knn5_nb <- knn2nb(knn5)
lw_knn <- nb2listw(knn5_nb, style = "W", zero.policy = TRUE)
```

## 8. Spatial Weights Summary

We printed a diagnostic summary of the KNN-5 spatial weights matrix,
confirming the number of spatial units and the distribution of neighbour
counts per unit. This verifies that the weight matrix was constructed
correctly before proceeding to autocorrelation tests.

```{r weights-summary, message=FALSE, warning=FALSE}
cat("\nSpatial weights created using KNN-5.\n")
cat("Number of units in spatial weights:", length(knn5_nb), "\n")
cat("Neighbors per unit:\n")
print(summary(card(knn5_nb)))
```

## 9. Global Moran's I Test (Asymptotic)

We tested for global spatial autocorrelation in TA-level teenage
pregnancy prevalence using the asymptotic Moran's I test, which assumes
that the standardised statistic follows a normal distribution under the
null hypothesis of no spatial clustering. The reported statistic,
expected value under the null, and p-value indicate whether the observed
spatial pattern is consistent with random geographical variation.

```{r moran-asymptotic, message=FALSE, warning=FALSE}
moran_asym <- moran.test(ta_lisa$prev, lw_knn, na.action = na.omit)
cat("Asymptotic Test:\n")
cat("  Moran's I:", round(moran_asym$statistic, 4), "\n")
cat("  Expected under null:", round(moran_asym$estimate[2], 4), "\n")
cat("  P-value:", round(moran_asym$p.value, 4), "\n\n")
```

## 10. Global Moran's I Permutation Test

We complemented the asymptotic test with a Monte Carlo permutation test
(999 simulations) that randomly reassigns prevalence values across TAs
to generate an empirical null distribution. The permutation p-value is
more robust when the normality assumption may not hold, providing a
cross-validation of the asymptotic result.

```{r moran-permutation, message=FALSE, warning=FALSE}
moran_perm <- moran.mc(ta_lisa$prev, lw_knn, nsim = 999, na.action = na.omit)
cat("Permutation Test, 999 simulations:\n")
cat("  Moran's I:", round(moran_perm$statistic, 4), "\n")
cat("  P-value:", round(moran_perm$p.value, 4), "\n")

if (moran_asym$p.value < 0.05) {
  cat("\n*** SIGNIFICANT SPATIAL CLUSTERING DETECTED, p < 0.05 ***\n")
} else {
  cat("\nNo significant spatial clustering detected, p >= 0.05.\n")
}
```

## 11. Local Moran's I / LISA Analysis

We computed Local Indicators of Spatial Association (LISA) using
`localmoran()` to identify individual TAs that contribute significantly
to the global autocorrelation. Each TA was classified into one of five
categories based on its quadrant position in the Moran scatter plot and
the local p-value (α = 0.05): High-High (hotspot), Low-Low (coldspot),
High-Low or Low-High (spatial outliers), and Not Significant.

```{r lisa-analysis, message=FALSE, warning=FALSE}
lmi <- localmoran(ta_lisa$prev, lw_knn, na.action = na.omit)

ta_lisa <- ta_lisa %>%
  mutate(
    Ii = lmi[, "Ii"],
    Z_Ii = lmi[, "Z.Ii"],
    p_lisa = lmi[, "Pr(z != E(Ii))"],
    lag_z = lag.listw(lw_knn, prev_z),
    cluster_type = case_when(
      is.na(prev_z)        ~ "No data",
      p_lisa >= 0.05       ~ "Not significant",
      prev_z > 0 & lag_z > 0 ~ "High-High (HOTSPOT)",
      prev_z < 0 & lag_z < 0 ~ "Low-Low (COLDSPOT)",
      prev_z > 0 & lag_z < 0 ~ "High-Low (Outlier)",
      prev_z < 0 & lag_z > 0 ~ "Low-High (Outlier)",
      TRUE ~ "Other"
    )
  )

print(table(ta_lisa$cluster_type))
```

## 12. Extract Hotspots

We extracted all High-High cluster TAs — areas with above-average
teenage pregnancy prevalence surrounded by neighbours that also have
above-average prevalence — and sorted them by prevalence in descending
order for reporting.

```{r hotspots, message=FALSE, warning=FALSE}
hotspots <- ta_lisa %>%
  filter(cluster_type == "High-High (HOTSPOT)") %>%
  arrange(desc(prev_pct)) %>%
  st_drop_geometry() %>%
  select(ta_name, n_teen, prev_pct, p_lisa, cluster_type)

if (nrow(hotspots) > 0) { print(hotspots) } else { cat("No significant hotspots detected.\n") }
```

## 13. Extract Coldspots

We extracted all Low-Low cluster TAs — areas with below-average teenage
pregnancy prevalence surrounded by similarly low-prevalence neighbours —
and sorted them by prevalence in ascending order. Coldspots indicate
geographic areas of relatively low burden that may serve as comparators
or reflect protective conditions.

```{r coldspots, message=FALSE, warning=FALSE}
coldspots <- ta_lisa %>%
  filter(cluster_type == "Low-Low (COLDSPOT)") %>%
  arrange(prev_pct) %>%
  st_drop_geometry() %>%
  select(ta_name, n_teen, prev_pct, p_lisa, cluster_type)

if (nrow(coldspots) > 0) { print(coldspots) } else { cat("No significant coldspots detected.\n") }
```

## 14. Hotspot Summary Table

We produced an interactive searchable summary table showing all
Traditional Authorities with their unweighted sample sizes, pregnant
counts, weighted prevalence, LISA p-values, and cluster classification.
Small-sample TAs are flagged to alert readers to interpret those
estimates with caution.

```{r hotspot-summary-table, message=FALSE, warning=FALSE}
ta_lisa %>%
  st_drop_geometry() %>%
  select(ta_name, n_teen, n_pregnant, prev_pct, p_lisa, cluster_type, flag_small) %>%
  arrange(desc(prev_pct)) %>%
  mutate(sample_size_flag = if_else(flag_small, "⚠", "✓")) %>%
  select(-flag_small) %>%
  gt() %>%
  tab_header(title = "Teen Pregnancy by Traditional Authority",
             subtitle = "Malawi MICS 2019–20: Survey-Weighted TA-Level LISA Analysis") %>%
  fmt_number(columns = c(prev_pct, p_lisa), decimals = 2) %>%
  fmt_number(columns = c(n_teen, n_pregnant), decimals = 0) %>%
  cols_label(ta_name = "Traditional Authority", n_teen = "N Teens",
             n_pregnant = "Pregnant", prev_pct = "Prevalence %",
             p_lisa = "LISA p-value", cluster_type = "Cluster Type",
             sample_size_flag = "Sample") %>%
  opt_interactive(use_search = TRUE, use_sorting = TRUE) %>%
  tab_options(table.font.size = px(11))
```

## 15. Moran Scatter Plot and LISA Map

We produced a two-panel figure combining a Moran scatter plot
(standardized prevalence vs. spatially lagged prevalence) and a LISA
choropleth map. The scatter plot annotates the global Moran's I
statistic and p-value and colors each TA by its LISA quadrant. The LISA
map displays statistically significant clusters (p \< 0.05) with
distinct colors for each LISA category, non-significant TAs in grey, and
district boundaries overlaid as dashed lines to provide geographic
context. A north arrow and scale bar are included.

```{r lisa-map, message=FALSE, warning=FALSE, fig.width=14, fig.height=7}

# Step 1: Compute quadrant classification for plotting
lisa <- localmoran(ta_lisa$prev, lw_knn, zero.policy = TRUE)

ta_lisa <- ta_lisa %>%
  mutate(
    local_I = lisa[, "Ii"],
    local_p = lisa[, "Pr(z != E(Ii))"],
    z_prev  = as.numeric(scale(prev)),
    z_lag   = as.numeric(scale(lag.listw(lw_knn, prev, zero.policy = TRUE))),
    quadrant = factor(case_when(
      z_prev > 0 & z_lag > 0 & local_p < 0.05 ~ "High-High (hotspot)",
      z_prev < 0 & z_lag < 0 & local_p < 0.05 ~ "Low-Low (coldspot)",
      z_prev > 0 & z_lag < 0 & local_p < 0.05 ~ "High-Low (outlier)",
      z_prev < 0 & z_lag > 0 & local_p < 0.05 ~ "Low-High (outlier)",
      TRUE ~ "Not significant"
    ), levels = c("High-High (hotspot)", "Low-Low (coldspot)",
                  "High-Low (outlier)", "Low-High (outlier)", "Not significant"))
  )

table(ta_lisa$quadrant)

# Step 2: Colors
lisa_cols <- c(
  "High-High (hotspot)" = "#E31A1C",
  "Low-Low (coldspot)"  = "#1F78B4",
  "High-Low (outlier)"  = "#FF7F00",
  "Low-High (outlier)"  = "#A6CEE3",
  "Not significant"     = "grey80"
)

# Step 3: Moran scatter plot
p_moran_scatter <- ggplot(st_drop_geometry(ta_lisa), aes(x = z_prev, y = z_lag)) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.4) +
  geom_vline(xintercept = 0, colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = quadrant, fill = quadrant), shape = 21, size = 2.4, stroke = 0.4, alpha = 0.85) +
  geom_smooth(method = "lm", se = TRUE, colour = "grey20", fill = "grey80", linewidth = 0.8) +
  annotate("text", x = Inf, y = -Inf,
    label = sprintf("Moran's I = %.3f\np = %.4f", moran_perm$statistic, moran_perm$p.value),
    hjust = 1.1, vjust = -0.5, size = 3.5, colour = "grey20", fontface = "italic") +
  annotate("text", x = Inf,  y = Inf,  label = "High-High", hjust = 1.1,  vjust = 1.5,  size = 3, colour = "#E31A1C", fontface = "bold") +
  annotate("text", x = -Inf, y = -Inf, label = "Low-Low",   hjust = -0.1, vjust = -0.5, size = 3, colour = "#1F78B4", fontface = "bold") +
  annotate("text", x = Inf,  y = -Inf, label = "High-Low",  hjust = 1.1,  vjust = -2.2, size = 3, colour = "#FF7F00", fontface = "bold") +
  annotate("text", x = -Inf, y = Inf,  label = "Low-High",  hjust = -0.1, vjust = 1.5,  size = 3, colour = "#A6CEE3", fontface = "bold") +
  scale_colour_manual(values = lisa_cols, guide = "none") +
  scale_fill_manual(values = lisa_cols, guide = "none") +
  labs(title = "Moran scatter plot",
       x = "Standardized teen pregnancy prevalence", y = "Spatially lagged prevalence") +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 12), panel.grid = element_blank())

## Extract district centroids for labeling (instead of TA labels)
district_labels <- districts %>%
  st_centroid() %>%
  mutate(district_name = GEONAMES)

# Then in the ggplot, replace the two geom_label_repel blocks with:

p_lisa_map <- ggplot() +
  # Base TA boundaries
  geom_sf(data = ta_boundaries_unique, fill = "grey96", colour = "grey75", linewidth = 0.25) +
  # Non-significant TAs in grey
  geom_sf(data = ta_lisa %>% filter(quadrant == "Not significant"),
          fill = "grey80", colour = "white", linewidth = 0.15) +
  # Significant TAs with LISA colors (main clusters)
  geom_sf(data = ta_lisa %>% filter(quadrant != "Not significant"),
          aes(fill = quadrant), colour = "white", linewidth = 0.2) +
  # District boundaries overlay - transparent fill with bold dark border
  geom_sf(data = districts, fill = NA, colour = "#000000", linewidth = 0.2, linetype = "solid", alpha = 0.7) +
  annotation_north_arrow(location = "tl", which_north = "true",
    height = unit(1.0, "cm"), width = unit(1.0, "cm"),
    pad_x = unit(0.2, "cm"), pad_y = unit(0.2, "cm"),
    style = north_arrow_fancy_orienteering(
      text_size = 7, text_face = "bold", line_width = 0.8,
      text_col = "grey20", fill = c("grey20", "white"))) +
  annotation_scale(location = "bl", width_hint = 0.30, text_cex = 0.65,
                   line_width = 0.6, height = unit(0.18, "cm"),
                   pad_x = unit(0.2, "cm"), style = "ticks") +
  scale_fill_manual(name = "LISA cluster type", values = lisa_cols, drop = FALSE) +
  labs(title = "LISA map: significant local spatial clusters") +
  coord_sf(datum = NA) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), axis.title = element_blank(),
        axis.text = element_blank(), axis.ticks = element_blank(),
        legend.position = "bottom", legend.title = element_text(face = "bold"),
        plot.title = element_text(face = "bold", size = 12)) +
  guides(fill = guide_legend(nrow = 2, ncol = 3, override.aes = list(linewidth = 0.4)))

# Step 5: Combine and display
p_lisa_combined <- p_moran_scatter + p_lisa_map +
  plot_annotation(
    title = "Spatial autocorrelation in teen pregnancy prevalence, Malawi MICS 2019–20",
    subtitle = sprintf("Left: Moran scatter plot (I = %.3f, p = %.4f). Right: LISA map at p < 0.05.",
                       moran_perm$statistic, moran_perm$p.value),
    caption = "k = 5 nearest neighbours, row-standardised. TA Nkula excluded. District boundaries (dashed lines) overlaid for geographic context. LISA p-values not corrected for multiple comparisons.",
    theme = theme(plot.title = element_text(face = "bold", size = 13, hjust = 0),
                  plot.subtitle = element_text(size = 10, colour = "grey35", hjust = 0),
                  plot.caption = element_text(size = 8, colour = "grey45", hjust = 0),
                  plot.margin = margin(8, 8, 6, 8))
  ) + plot_layout(widths = c(1, 1))

p_lisa_combined
```

## 16. Save LISA Figure

We exported the combined LISA figure in two formats: a high-resolution
PNG (300 dpi) for manuscript submission and a TIFF (600 dpi, LZW
compression) for journal requirements. Both files are saved to the
working directory.

```{r save-figure, message=FALSE, warning=FALSE}
ggsave(
  "Figure_LISA_Teen_Pregnancy_Malawi_MICS.png",
  plot = p_lisa_combined,
  dpi = 300,
  width = 14,
  height = 7,
  units = "in",
  bg = "white"
)

ggsave(
  "Figure_LISA_Teen_Pregnancy_Malawi_MICS.tiff",
  plot = p_lisa_combined,
  device = "tiff",
  dpi = 600,
  width = 14,
  height = 7,
  units = "in",
  compression = "lzw"
)
```

## 17. Export Spatial Results

We exported the final `ta_lisa` spatial object as a shapefile for use in
external GIS software (ArcGIS or QGIS), enabling further cartographic
styling, overlay analysis, or reporting. The exported file includes all
LISA statistics, cluster classifications, prevalence estimates, and
sample size flags.

```{r export-results, message=FALSE, warning=FALSE}
cat("\n", strrep("=", 80), "\n", sep = "")

cat("EXPORTING RESULTS\n")

cat(strrep("=", 80), "\n", sep = "")

st_write(
  ta_lisa,
  "ta_lisa_results_with_boundaries.shp",
  delete_dsn = TRUE
)

cat("Exported: ta_lisa_results_with_boundaries.shp\n")

cat("Use this in ArcGIS or QGIS to map hotspots, coldspots, and prevalence.\n")
```

## 18. Export CSV Results

We exported the non-spatial LISA results as a flat CSV file for use in
supplementary tables, secondary analysis, or sharing with co-authors who
may not work in R or GIS environments.

```{r export-csv, message=FALSE, warning=FALSE}
write_csv(
  ta_lisa %>% st_drop_geometry(),
  "ta_lisa_results.csv"
)

cat("Exported: ta_lisa_results.csv\n")
```

## 19. Export Summary Statistics

We wrote a plain-text summary report containing the full global Moran's
I test output (asymptotic and permutation), the LISA cluster
classification counts, and the hotspot and coldspot TA lists. This file
serves as a reproducible record of all key spatial statistics produced
by the analysis.

```{r export-summary-stats, message=FALSE, warning=FALSE}
sink("ta_lisa_summary_statistics.txt")

cat("MALAWI MICS 2019–20: TA-LEVEL LISA ANALYSIS OF TEEN PREGNANCY\n")
cat(strrep("=", 80), "\n\n", sep = "")

cat("GLOBAL MORAN'S I TEST:\n")
print(moran_asym)

cat("\n\nPERMUTATION TEST, 999 simulations:\n")
print(moran_perm)

cat("\n\nCLUSTER CLASSIFICATION:\n")
print(table(ta_lisa$cluster_type))

cat("\n\nHOTSPOTS, High-High:\n")
if (nrow(hotspots) > 0) {
  print(hotspots)
} else {
  cat("None detected\n")
}

cat("\n\nCOLDSPOTS, Low-Low:\n")
if (nrow(coldspots) > 0) {
  print(coldspots)
} else {
  cat("None detected\n")
}

sink()

cat("Exported: ta_lisa_summary_statistics.txt\n")
```

# PART V: Discussion and Interpretation

## Key Findings: Teenage Pregnancy in Malawi

### Prevalence and Sociodemographic Risk Factors

Among the 5,758 adolescent females (weighted n = 5,538) examined in this
analysis, the overall prevalence of teenage pregnancy was 25.5%.
Substantial sociodemographic disparities were evident in the descriptive
analysis (Table 2), with prevalence ranging from 10.3% among the richest
households to 42.8% among the poorest—a 4.2-fold difference. Education
was similarly patterned: girls with primary education or lower had a
prevalence of 31.5%, compared to 14.7% among those with secondary
education or higher, reflecting 2.1-fold disparity. Rural residence
carried 1.9 times the prevalence of urban residence (29.6% vs. 15.3%).

The strongest univariable associations with pregnancy (Table 3, Model 1)
were current marital status (OR 51.1 for currently married vs. not in
union, 95% CI 38.5–67.8), contraceptive use (OR 11.3, 95% CI 7.45–17.1),
and wealth index (OR 6.50 for poorest vs. richest, 95% CI 4.86–8.69).
These patterns align with demographic expectations: marriage and union
formation precede and facilitate childbearing among adolescents; current
contraceptive use is a strong marker of prior reproductive experience;
and economic hardship (poverty) is a documented risk factor for early
pregnancy in low-income settings.

### Adjusted Risk Factor Model

In the multivariable model adjusted for all six sociodemographic
exposures (Table 3, Model 2, n = 4,776 unweighted; n = 4,633 weighted),
the adjusted odds ratios (aORs) revealed several important patterns:

1.  **Wealth gradient persisted:** Compared to the richest quintile,
    girls in lower wealth quintiles retained significantly elevated
    risk, with aORs ranging from 2.71 (second quintile, 95% CI
    1.45–5.05) to 2.87 (poorest, 95% CI 1.55–5.32). The gradient was
    remarkably consistent across the four lower quintiles, confirming
    that poverty-related factors are robust risk drivers even after
    accounting for education, residence, marital status, contraception,
    and insurance.

2.  **Marital status dominated:** Marriage or co-habitation remained the
    strongest predictor after adjustment, with aOR 35.3 (95% CI
    24.9–50.0) for currently married girls and aOR 40.7 (95% CI
    17.9–92.5) for those living with a partner. This reflects the
    biological reality that marriageability and childbearing are tightly
    coupled in this setting and that married adolescents are nearly
    certain to be pregnant or to become pregnant within the marriage.

3.  **Education lost statistical significance after adjustment** (aOR
    0.77 for secondary or higher vs. primary or lower, 95% CI
    0.56–1.06). This suggests that much of education's protective effect
    in the crude analysis operates through its correlation with wealth,
    marital status, or other mediators—a finding worth exploring in
    future mediation analyses.

4.  **Place of residence also became non-significant** (aOR 0.68 for
    rural vs. urban, 95% CI 0.40–1.17), implying that the rural–urban
    disparity is largely confounded by differences in wealth, marital
    patterns, and education across these regions.

5.  **Contraceptive use remained strongly associated with pregnancy**
    (aOR 11.7, 95% CI 7.02–19.5, p \< 0.001). However, this association
    is most naturally interpreted as **reverse causality**: girls who
    are pregnant or have experienced pregnancy are more likely to adopt
    contraception to prevent future pregnancy. This variable marks
    reproductive history rather than a causal risk factor and should not
    be interpreted as "using contraception increases pregnancy risk."
    Its inclusion in the multivariable model provides adjustment for
    reproductive experience among the unpregnant (those not yet pregnant
    or no longer pregnant at survey), but causal interpretation is
    unwarranted.

6.  **Health insurance had insufficient precision** to inform policy:
    only 1 pregnant case was observed among 22 weighted insured
    adolescents (aOR 0.72, 95% CI 0.11–4.75). The wide confidence
    interval encompasses both protection and increased risk. This
    variable adds little information and reflects both the low insurance
    coverage in the survey and its weak association with pregnancy
    prevalence in this population. Results should not be interpreted as
    evidence of insurance protecting against or increasing pregnancy
    risk.

### Spatial Clustering and Geographic Hotspots

Global spatial autocorrelation analysis revealed **statistically
significant clustering** of teenage pregnancy at the Traditional
Authority level. Both the asymptotic Moran's I test (I = 2.998, p =
0.0014) and the permutation test with 999 simulations (I = 0.094, p =
0.002) confirmed that TAs with high (or low) prevalence are surrounded
by geographically similar TAs—contrary to random spatial variation. This
clustering signal justifies and validates the spatial analysis approach.

Local Indicators of Spatial Association (LISA) classified the 331
surveyed TAs into five categories: - **6 hotspots** (High-High,
spatially significant clusters of above-average prevalence) - **21
coldspots** (Low-Low, spatially significant clusters of below-average
prevalence) - **10 outliers** (7 High-Low; 3 Low-High) - **294 not
significant**

The six hotspots identified were: 1. **Ngabu Urban** (71.4% prevalence,
n = 7) — located in Lower Shire, Southern Region 2. **TA Changata**
(56.3%, n = 16) — Lower Shire 3. **TA Sitola** (42.9%, n = 7) — Lower
Shire 4. **TA Mbenje** (42.6%, n = 28) — Lower Shire 5. **TA Ngabu**
(38.0%, n = 69) — Lower Shire (larger sample) 6. **Mwabvi Game Reserve**
(28.6%, n = 7) — Lower Shire

These six TAs form a **coherent geographic cluster in the Lower Shire
region** of southern Malawi, suggesting localized drivers of high
teenage pregnancy (e.g., regional differences in girls' education
access, marriage norms, economic opportunities, health service
availability, or social norms around early childbearing). Further
investigation into district-level interventions, service gaps, and
community-level risk and protective factors in the Lower Shire is
warranted.

**Caveat on hotspot sample sizes:** Three of the six hotspots (Ngabu
Urban, TA Sitola, Mwabvi Game Reserve) have unweighted samples of only 7
girls each. While the LISA test is statistically significant (p \< 0.05
for each), these estimates are based on very small samples and should be
interpreted with appropriate caution. The two coldspots with zero
prevalence (Area 21, Area 22, etc.) are similarly small-sample urban
wards; their classification as coldspots is statistically valid but may
reflect chance variation or geographic concentration of more
economically advantaged populations in these urban areas rather than
deep structural protection from pregnancy.

### Spatial Heterogeneity and Implications for Policy

The identification of a high-prevalence cluster in the Lower Shire
suggests that **one-size-fits-all national approaches to adolescent
sexual and reproductive health may be insufficient**. Targeted
interventions—including enhanced reproductive health education, improved
contraceptive access, economic opportunity expansion for girls, and
marriage postponement programming—may be particularly impactful if
concentrated in high-prevalence TAs and neighboring regions.

Conversely, coldspots (predominantly urban administrative units with
zero or near-zero prevalence in the survey) may offer comparative
examples of protective factors to study and potentially scale.

### Limitations

1.  **Small sample sizes in spatial units:** Many TAs had fewer than 20
    teen respondents (some fewer than 5), limiting the precision of
    prevalence estimates and the stability of LISA classifications.
    Estimates for very-small-sample TAs should be interpreted as
    preliminary or flagged for future rounds of larger-scale
    surveillance.

2.  **Cross-sectional design:** This analysis cannot establish
    causality. The associations observed in Table 3, particularly for
    contraceptive use, are subject to reverse causality and other
    confounding not measured in the MICS (e.g., age at sexual debut,
    partner age, access to family planning, community-level norms).

3.  **GPS jittering:** Cluster coordinates in the MICS are jittered (+/-
    2–10 km urban, +/- 5 km rural) for confidentiality, which can
    introduce noise into spatial analyses. TAs with few clusters may be
    more sensitive to this jitter.

4.  **Missing data:** While item-level missingness was low (\<2%), six
    girls (0.1%) were excluded due to missing age or pregnancy status.
    Complete-case analysis in the multivariable model (n = 4,776)
    excludes girls with missing data on any covariate; patterns of
    missingness by wealth or education were not formally examined.

5.  **Unmeasured confounding:** The MICS does not include information on
    girls' age at sexual debut, partner age, access to and quality of
    reproductive health services, community and religious norms about
    early marriage and childbearing, or intergenerational transmission
    of early pregnancy. These factors may be important mediators or
    confounders.

6.  **Health insurance prevalence:** Only 0.4% of adolescent respondents
    reported health insurance, limiting statistical power and
    generalizability of insurance-related estimates.

### Strengths

-   **Complex survey design properly accounted for** via two-stage
    stratified cluster sampling specification in the `survey` package,
    ensuring estimates are representative of the adolescent population
    aged 15–19 in Malawi.
-   **Survey-weighted and internally consistent analysis:** All
    prevalence, regression, and spatial analyses used the same weighting
    scheme and complete-case sample, avoiding discrepancies between
    tables.
-   **Convergent spatial evidence:** Both asymptotic and permutation
    Moran's I tests confirm clustering, increasing confidence in the
    spatial signal.
-   **Transparent reference categories and model specification:** All
    reference categories are pre-specified, reported, and held constant
    across tables, facilitating reproducibility and policy
    communication.

## 20. Save R Workspace

We saved the complete R workspace to an `.RData` file so that all
objects — spatial weights, survey designs, prevalence estimates, and
LISA results — can be reloaded in future sessions without re-running the
full analysis pipeline.

```{r save-rdata, message=FALSE, warning=FALSE}
save.image("ta_lisa_analysis_with_boundaries.RData")

cat("Saved: ta_lisa_analysis_with_boundaries.RData\n")
```

## 21. Analysis Complete

This chunk prints a completion message and a checklist of recommended
next steps for post-processing and reporting. All analytical outputs —
tables, figures, shapefiles, and summary statistics — are saved to the
working directory.

```{r analysis-complete, message=FALSE, warning=FALSE}
cat("\n", strrep("=", 80), "\n", sep = "")

cat("ANALYSIS COMPLETE\n")

cat(strrep("=", 80), "\n", sep = "")

cat("\nNext steps:\n")

cat("1. Open ArcGIS or QGIS.\n")

cat("2. Load ta_lisa_results_with_boundaries.shp.\n")

cat("3. Style by cluster_type to visualize hotspots and coldspots.\n")

cat("4. Or style by prev_pct to show the prevalence gradient.\n")
```
